The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\*[-8pt] for the free eLearning program CrypTool (version 1.4.10 Beta 4-Pre3)} \vspace{60pt}

### 2007-05-11 script/en/chapters/numbertheory.tex
Script: bugfixes and updates like M44 and Sierpinki

\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table 5 for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table 6 demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$ are of
    order 22.
  } For a modulus $n$ and $a$ relative prime to
  $n$ the following holds: \\
  $\{ a^i (\mbox{mod }n) | i = 1,\dots,J(n)\}$ equals the multiplicative
  group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}

@@

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice)\index{Mathematica}:
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\index{Pari-GP}\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.
}${}^,$\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\index{Mathematica}\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.\index{Pari-GP}
\vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +6 pt
} is extremely difficult to solve.

@@

   \item BC, \index{BC} \\
          Interpreter with number-theory functions \\
          \href{http://www.maths.uq.edu.au/~krm/gnubc.html}
          {\texttt{http://www.maths.uq.edu.au/\~{}}krm/gnubc.html}

@@

    \item Once again only after finishing this I happened upon
          the web site of Mr.~Wagner, who is responsible for the development
	  of the curriculum of computer science in one of the German federal 
          states (L\"ander). Here you can get hold of a collection of texts
	  and (Java-)\discretionary{}{}{}programs (available only in German):\\
          \href{http://www.hom.saar.de/~awa/kryptolo.htm}
          {\texttt{http://www.hom.saar.de/\~{}awa/kryptolo.htm}}

>>>>

\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table 5 for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table 6 demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  } Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}
The multiplicative group $Z_n^*$ contains all values from $1$ to $n-1$.

@@

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice)\index{Mathematica}:
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\index{Pari-GP}\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or \index{BC} BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.
}${}^,$\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\index{Mathematica}\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.\index{Pari-GP}
\vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +6 pt
} is extremely difficult to solve.

@@

   \item BC, \index{BC} \\
          Interpreter with number-theory functions \\
         \href{http://www.gnu.org/software/bc/bc.html}
         {\texttt{http://www.gnu.org/software/bc/bc.html}}

@@

    \item Web site of Mr.~Wagner, who is responsible for the development
	  of the curriculum of computer science in one of the German federal 
          states (L\"ander). Here you can get hold of a collection of texts
	  and (Java-)\discretionary{}{}{}programs (available only in German):\\
          \href{http://www.hom.saar.de/~awa/kryptolo.htm}
          {\texttt{http://www.hom.saar.de/\~{}awa/kryptolo.htm}}

### 2007-05-11 script/en/chapters/primes.tex
Script: bugfixes and updates like M44 and Sierpinki

\newpage
\hypertarget{Kapitel_2}{}
\section{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006)

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsubsection{The 11 largest known primes (as of March 2006)} 
\label{RecordPrimes}
\index{Prime number!records}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline
	1  & $2^{30,402,457}-1$ & 9,152,052 & 2005 & Mersenne, 43rd known \\
	2  & $2^{25,964,951}-1$ & 7,816,230 & 2005 & Mersenne, 42nd known \\
	3  & $2^{24,036,583}-1$ & 7,235,733 & 2004 & Mersenne, 41st known \\
	4  & $2^{20,996,011}-1$ & 6,320,430 & 2003 & Mersenne, 40th known \\
	5  & $2^{13,466,917}-1$ & 4,053,946 & 2001 & Mersenne, M-39 \\
	6  & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\
	7  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	8  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
	9  & $2^{ 3,021,377}-1$ &   909,526 & 2001 & Mersenne, M-37 \\
	10  & $2^{ 2,976,221}-1$ &   895,932 & 2001 & Mersenne, M-36 \\
	11 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\

\hline
\end{tabular}
\caption{The 11 largest known primes and its particular number types
         (as of March 2006)}
\end{center}
\end{table} 
\footnotetext{$ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $ } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

The largest known prime is a Mersenne prime.
This prime was found by the \hyperlink{GIMPS-project}{GIMPS project}
(chapter~\ref{zahlentyp_mersenne}).

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \href{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}
   {\texttt{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}}. \\
Also see:
      \href{http://www.utm.edu/}
   {\texttt{http://www.utm.edu/}}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221;
 ~ 3,021,377; ~ 6,972,593; \\
 ~ 13,466,917;  ~ 20,996,011;  ~ 24,036,583;  ~ 25,964,951;  ~ 30,402,457.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$43$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \href{http://www.mersenne.org/status.htm}
   {\texttt{http://www.mersenne.org/status.htm}}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}.

                                                 }.
The $19$th number with the exponent $4,253$ was the first with at least $1,000$ digits in  decimal system
(the mathematician Samual \index{Yates, Samual} Yates coined the expression {\em
titanic} \index{Prime number!titanic} prime for this; it was discovered by
Hurwitz in 1961); the $27$th number with the exponent $44,497$ was the first
with at least $10,000$ digits in the decimal system (Yates coined the expression
\index{Prime number!gigantic}  {\em gigantic} prime for this. These names are
now long outdated).

@@

Right now the GIMPS project has discovered nine
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
largest Mersenne primes so far, including the largest known prime number at all.

@@

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[h]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline
	$2^{30,402,457}-1$ & 9,152,052 & December 15, 2005 & Curtis Cooper/Steven Boone \\
	$2^{25,964,951}-1$ & 7,816,230 & February 18, 2005 & Martin Nowak     \\
	$2^{24,036,583}-1$ & 7,235,733 & May 15, 2004      & Josh Findley     \\
	$2^{20,996,011}-1$ & 6,320,430 & November 17, 2003 & Michael Shafer   \\
	$2^{13,466,917}-1$ & 4,053,946 & November 14, 2001 & Michael Cameron  \\
	$2^{ 6,972,593}-1$ & 2,098,960 & June 1, 1999      & Nayan Hajratwala \\
	$2^{ 3,021,377}-1$ &   909,526 & January 27, 1998  & Roland Clarkson  \\
	$2^{ 2,976,221}-1$ &   895,932 & August 24, 1997   & Gordon Spence    \\
	$2^{ 1,398,269}-1$ &   420,921 & November 1996     & Joel Armengaud   \\

\hline
\end{tabular}
\caption{The largest primes found by the GIMPS project (as of March 2006)}
\end{center}
\end{table}

@@

>>>>

\newpage
\hypertarget{Kapitel_2}{}
\section{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, May 2007)

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsubsection{The 20 largest known primes (as of May 2007)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{Prime number!records}

@@

\index{Mersenne!number!generalized}
\index{Fermat!number!generalized}

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline
	1  & $2^{32,582,657}-1$ & 9,808,358 & 2006 & Mersenne, 44th known \\
	2  & $2^{30,402,457}-1$ & 9,152,052 & 2005 & Mersenne, 43rd known \\
	3  & $2^{25,964,951}-1$ & 7,816,230 & 2005 & Mersenne, 42nd known \\
	4  & $2^{24,036,583}-1$ & 7,235,733 & 2004 & Mersenne, 41st known \\
	5  & $2^{20,996,011}-1$ & 6,320,430 & 2003 & Mersenne, 40th known \\
	6  & $2^{13,466,917}-1$ & 4,053,946 & 2001 & Mersenne, M-39 \\
	7  & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\
	8  & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\

	9  & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\

	10  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	11  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\

	12  & $4,847 \cdot 2^{3,321,063}+1$ & 999,744 & 2005 & Generalized Mersenne \\
	13  & $3 \cdot 2^{3,136,255}-1$ & 944,108 & 2007 & Generalized Mersenne \\

	14  & $2^{ 3,021,377}-1$ &   909,526 & 1998 & Mersenne, M-37 \\
	15  & $2^{ 2,976,221}-1$ &   895,932 & 1997 & Mersenne, M-36 \\

	16  & $222,361 \cdot 2^{2,854,840}+1$ & 859,398 & 2006 & Generalized Mersenne \\

	17 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\

	18  & $1,361,244^{131,072}+1$ &   803,988 & 2004 & Generalized Fermat \\
	19  & $1,176,694^{131,072}+1$ &   795,695 & 2003 & Generalized Fermat \\

	20  & $342,673 \cdot 2^{2,639,439}-1$ & 794,556 & 2007 & Generalized Mersenne \\

\hline
\end{tabular}
\caption{The 20 largest known primes and its particular number types
         (as of May 2007)}    % Eyecatcher_neue-Mersenne
\end{center}
\end{table} 
\footnotetext{\index{Number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\href{http://www.seventeenorbust.com}{\texttt{http://www.seventeenorbust.com}})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its
task sometime.

The SoB project tries to prove computationally, that the number $k = 78,557$ is
the smallest Sierpinski number (John Selfridge proved in 1962, that $78,557$ is a Sierpinski number).

The famous Polish mathematician Waclaw Sierpinski (1882 to 1969) proved in
1960, that there exist infinitely many odd integers k, which fulfill the
following property: For all Sierpinski numbers k it is true: All numbers $N = k \cdot 2^{n}+1$ are composite for all integers $n>=1$ (Sierpinski's Composite Number Theorem, \href{http://mathworld.wolfram.com/SierpinskisCompositeNumberTheorem.html}{\texttt{http://mathworld.wolfram.com/SierpinskisCompositeNumberTheorem.html}}).

When the project started in 2002 there have been 17 possible candidates $< 78557$ (this is the reason for the project's name ``Seventeen or Bust''). It is sufficient to find one single counter-example, to exclude a candidate k, which means to find a single $n>=1$, where $N = k \cdot 2^{n}+1$ is prime. So it is only a byproduct of this task that this also generates new monster primes.
} 
\footnotetext{%
Generalized Fermat number: $ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $
\index{Fermat!number!generalized}  } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

The largest known prime is a Mersenne prime, found by the
\hyperlink{GIMPS-project}{GIMPS project}(chapter~\ref{zahlentyp_mersenne}).

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \href{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}
   {\texttt{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}}. \\
Also see:
      \href{http://www.utm.edu/}
   {\texttt{http://www.utm.edu/}}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221;
 ~ 3,021,377; ~ 6,972,593; \\
 ~ 13,466,917;  ~ 20,996,011;  ~ 24,036,583;  ~ 25,964,951;  ~ 30,402,457,  ~ 32.582.657.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$44$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}.

The $19$th number with the exponent $4,253$ was the first with at least $1,000$ digits in  decimal system
(the mathematician Samual \index{Yates, Samual} Yates coined the expression {\em
titanic} \index{Prime number!titanic} prime for this; it was discovered by
Hurwitz in 1961); the $27$th number with the exponent $44,497$ was the first
with at least $10,000$ digits in the decimal system (Yates coined the expression
\index{Prime number!gigantic}  {\em gigantic} prime for this. These names are
now long outdated).

\vskip +25 pt
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \href{http://www.mersenne.org/status.htm}
   {\texttt{http://www.mersenne.org/status.htm}}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}. }.

@@

Right now the GIMPS project has discovered ten
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
largest Mersenne primes so far, including the largest known prime number at all.

@@

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[h]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline
	$2^{32,582,657}-1$ & 9,808,358 & September 4, 2006 & Curtis Cooper/Steven Boone \\
	$2^{30,402,457}-1$ & 9,152,052 & December 15, 2005 & Curtis Cooper/Steven Boone \\
	$2^{25,964,951}-1$ & 7,816,230 & February 18, 2005 & Martin Nowak     \\
	$2^{24,036,583}-1$ & 7,235,733 & May 15, 2004      & Josh Findley     \\
	$2^{20,996,011}-1$ & 6,320,430 & November 17, 2003 & Michael Shafer   \\
	$2^{13,466,917}-1$ & 4,053,946 & November 14, 2001 & Michael Cameron  \\
	$2^{ 6,972,593}-1$ & 2,098,960 & June 1, 1999      & Nayan Hajratwala \\
	$2^{ 3,021,377}-1$ &   909,526 & January 27, 1998  & Roland Clarkson  \\
	$2^{ 2,976,221}-1$ &   895,932 & August 24, 1997   & Gordon Spence    \\
	$2^{ 1,398,269}-1$ &   420,921 & November 1996     & Joel Armengaud   \\

\hline
\end{tabular}
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\caption{The largest primes found by the GIMPS project (as of May 2007)}
\end{center}
\end{table}

@@

\bibitem[Shoup2005]{Shoup2005} \index{Shoup 2005}
    Victor Shoup \\
    {\em A Computational Introduction to Number Theory and Algebra},\\
    Cambridge University Press, 2005, \\
    \href{http://shoup.net/ntb/}{\texttt{http://shoup.net/ntb/}}.
